{\bf {\Large Improving estimates of ecosystem metabolism by reducing effects of tidal advection on dissolved oxygen time series}}

### 2015-05-29 mat_proc.Rnw
finished journal revisions to reviewers after co-author suggestions

\subsection{Weighted regression for modelling and filtering \ac{DO} time series}

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations.  The beginning of each day was defined as the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method, which included parameters to estimate variation of the response variable on a sinusoidal period.  \ac{DO} variation was not modeled using this approach because the expected metabolically-driven pattern is not strictly sinusoidal.  Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  Moreover, tidal height measurements are more widely available and are generally monitored by water quality sondes.

Weighted regression was implemented as a moving window that allowed for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models were estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantified the relevance of observations to the center of the window in respect to time, hour of the day, and tidal height.  Specifically, weights were assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher weight in the regression.  Observations that exceed the maximum width of the window $h$ are assigned a weight of zero.  The tri-cube weighting function is similar to a Gaussian distribution such that weights decrease gradually from the center until the maximum window width is reached.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of the three separate weight vectors for time (day), hour, and tidal height. Windows for time and hour are used to weight observations based on distance (time) from the center of the window.  The window for tidal height is used to weight observations based on the difference from mean tide as a proportion of tide range.  For example, a half-window width of 0.5 means that observations are weighted proportionately within +/- 50\% of the tide range referenced to the tidal height in the center of the window. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

>>>>

\subsection{Weighted regression for modelling and filtering \ac{DO} time series}\label{mat_proc}

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations.  The beginning of each day was defined as the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method, which included parameters to estimate variation of the response variable on a sinusoidal period.  \ac{DO} variation was not modeled using this approach because the expected metabolically-driven pattern is not strictly sinusoidal.

Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  An important distinction between tidal height and advection is that each variable could provide different information about tidal regime.  For example, tidal heights at the minimum and maximum of the range for standing wave tides may be associated with periods of low advection when water masses are not moving rapidly past the sensor.  The relationship between tidal height and advection across multiple environments may be complex.  The weighted regression model is suited for application to different locations because it needs only a variable that indicates a particular point in the tidal cycle.  The specific influence of advection on the \ac{DO} time series is isolated using the moving window approach such that the relationship between tidal height and advection does not need to be known.

Weighted regression was implemented as a moving window that allowed for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models were estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantified the relevance of observations to the center of the window in respect to time, hour of the day, and tidal height.  Specifically, weights were assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher weight in the regression.  Observations that exceed the maximum width of the window $h$ are assigned a weight of zero.  The tri-cube weighting function is similar to a Gaussian distribution such that weights decrease gradually from the center until the maximum window width is reached.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of the three separate weight vectors for time (day), hour, and tidal height. Windows for time (continuous throughout the day) and hour (within each day) are used to weight observations based on distance (time) from the center of the window.  The window for tidal height is used to weight observations based on the difference from mean tide as a proportion of tide range.  For example, a half-window width of 0.5 means that observations are weighted proportionately within +/- 50\% of the tide range referenced to the tidal height in the center of the window. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

### 2015-05-29 swmp_noise.tex
finished journal revisions to reviewers after co-author suggestions

In aquatic ecosystems, time series of \ac{DO} have been used to compute estimates of ecosystem metabolism.  Central to this open water or `Odum' method is the assumption that the dissolved oxygen time series is a Lagrangian specification of the flow field.  However, most \ac{DO} time series are collected at fixed locations, such that changes in dissolved oxygen are assumed to reflect metabolism and that effects of advection or mixing can be neglected.  A statistical model using weighted regression was applied to separate variability in \ac{DO} from tidal variation or other advection in estuaries, thereby helping to partially relax this assumption and improve metabolism estimates. The method targets the periodicity of the tidal component while preserving the true biological signal, offering a distinct advantage over traditional deconvulution methods.  The model was developed and tested using simulated \ac{DO} time series with known biological and physical components, and then applied to one year of continuous monitoring data from four stations within the \acl*{NERRS}.  Variability from advection was reduced on average by 70.2\% for production and 74.3\% for respiration.  The model was especially effective when the magnitude of tidal influence was high and correlations between tidal change and the solar cycle were low, whereas collinearity in the latter instance limited model performance. By reducing the effects of physical transport on metabolism estimates, there may be increased potential to empirically relate metabolic rates to causal factors on timescales of several days to several weeks.

@@

\subsection{Weighted regression for modelling and filtering \ac{DO} time series}

@@

Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  An important distinction between tidal height and advection is that each variable could provide different information about a tidal regime.  Tidal heights at the minimum and maximum of the range may be associated with periods of low advection when water masses are not moving rapidly past the sensor, whereas tidal heights near the mean may be more likely to have greater advection.  Accordingly, our use of tidal height should not be confused with a variable that directly measures advection.  The model only requires a variable that indicates a particular point in the tidal cycle, such that tidal height can be mapped to advection with quantifiable periodicity that can be isolated.

@@

The weighted regression method was applied to the annual data for each station to obtain a filtered \ac{DO} time series for estimating metabolism.  Ecosystem metabolism was estimated using the open-water technique \citep{Odum56} as described in \citet{Caffrey14}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey14}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO} concentration, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  The diffusion-corrected \ac{DO} flux estimates as hourly rates of \ac{DO} change were first averaged during day and night periods for each 24 hour `metabolic day' in the time series. The `metabolic day' was considered the approximate 24 hour period between sunsets on two adjacent calendar days.  Hourly \ac{DO} flux was considered respiration during night hours and net production during day hours.  Total respiration (Rt) rates were assumed constant during day and night such that daily rates were calculated as the average \ac{DO} flux during night hours multiplied by 24. Daily gross production (Pg) was the average \ac{DO} flux during day hours minus the average \ac{DO} flux during night hours, multiplied by total sunlight time.  Net ecosystem metabolism was gross production (positive) plus total respiration (negative).  Finally, volumetric rates were converted to depth-integrated (areal) estimates (g O$_2$ m$^{-2}$ d$^{-1}$, \cref{tab:case_att}) by multiplying by the mean water-column depth from the water quality sensor.  A half-meter was also added to account for approximate placement of the sensor slightly off of the bottom.

@@

For all case studies, weighted regression was generally successful in reducing the variation in the \ac{DO} time series that was presumably caused by physical advection.  In particular, filtering reduced the frequency of anomalous metabolism estimates, such as negative production and positive respiration \citep{Needoba12}.  It is important to recognize, however, that `normal' estimates (positive production and negative respiration) may still include significant bias due to physical advection.  For example, a large decrease in \ac{DO} at night may appear `normal' even if inflated by a tidal effect.  In some cases, apparent tidal effects could reflect real ecosystem processes. For example, \citet{Nidzieko14} observed that net metabolism at Elkhorn Slough was strongly heterotrophic during spring tides that occurred at nighttime because inundation of salt marshes during the night followed by draining during a falling tide during the day lead to larger respiration values.  Resolving these patterns may simply require better data.  For the weighted regression approach, sustained periods of synchrony between the solar and tidal cycles presents a problem because collinearity between the two diminishes performance.  Moreover, \ac{DO} time series may reflect changes in metabolism associated with tides that are true biological effects, rather than an artifact of advection.  For example, respiration may be higher on falling tides as compared to rising tides if outwelling substrates on the falling tide support increased metabolism \citep{Sasaki09}.  Although one might hope that an improved ability to resolve temporal patterns in metabolism would make it possible to resolve such changes using only a \ac{DO} time series, we suggest that this is still not possible.  The open water metabolism method has generally been applied at the time scale of one day, rather than hourly.  Therefore, weighted regression may improve estimates of daily metabolism, but will not make it possible to resolve intraday variations in metabolism.

>>>>

In aquatic ecosystems, time series of \ac{DO} have been used to compute estimates of ecosystem metabolism.  Central to this open water or `Odum' method is the assumption that the dissolved oxygen time series is a Lagrangian specification of the flow field.  However, most \ac{DO} time series are collected at fixed locations, such that changes in dissolved oxygen are assumed to reflect metabolism and that effects of advection or mixing can be neglected.  A statistical model using weighted regression was applied to separate variability in \ac{DO} from tidal variation or other advection in estuaries, thereby helping to partially relax this assumption and improve metabolism estimates. The method explicitly targets the tidal component using a flexible association while preserving the true biological signal, offering a distinct advantage over traditional deconvulution methods. The model was developed and tested using simulated \ac{DO} time series with known biological and physical components, and then applied to one year of continuous monitoring data from four stations within the \acl*{NERRS}.  Variability from advection was reduced in all cases, in addition to a reduction of the correlation with the tide.  The model was especially effective when the magnitude of tidal influence was high and correlations between tidal change and the solar cycle were low, whereas collinearity in the latter instance limited model performance. By reducing the effects of physical transport on metabolism estimates, there may be increased potential to empirically relate metabolic rates to causal factors on timescales of several days to several weeks.

@@

\subsection{Weighted regression for modelling and filtering \ac{DO} time series}\label{mat_proc}

@@

Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  An important distinction between tidal height and advection is that each variable could provide different information about tidal regime.  For example, tidal heights at the minimum and maximum of the range for standing wave tides may be associated with periods of low advection when water masses are not moving rapidly past the sensor.  The relationship between tidal height and advection across multiple environments may be complex.  The weighted regression model is suited for application to different locations because it needs only a variable that indicates a particular point in the tidal cycle.  The specific influence of advection on the \ac{DO} time series is isolated using the moving window approach such that the relationship between tidal height and advection does not need to be known.

@@

The weighted regression method was applied to the annual data for each station to obtain a filtered \ac{DO} time series for estimating metabolism.  Ecosystem metabolism was estimated using the open-water technique \citep{Odum56} as described in \citet{Caffrey14}.  The method is used to infer net ecosystem metabolism using the mass balance equation:
\begin{equation} \label{metrate}
\frac{\delta DO}{\delta t} = P - R + D
\end{equation}
where the change in \ac{DO} concentration ($\delta DO$, g O$_2$ m$^{-3}$) over time ($\delta t$, hours) is equal to photosynthetic rate ($P$, g O$_2$ m$^{-3}$ hr$^{-1}$) minus respiration rate ($R$, g O$_2$ m$^{-3}$ hr$^{-1}$), corrected for air-sea gas exchange ($D$, g O$_2$ m$^{-3}$ hr$^{-1}$) \citep{Caffrey14}. $D$ is estimated as the difference between the \ac{DO} saturation concentration and observed \ac{DO} concentration, multiplied by a volumetric reaeration coefficient, $k_a$ \citep{Thebault08}.  This coefficient accounts for the influence of meteorological effects, such as wind, on air-sea gas exchange at the surface.  The diffusion-corrected \ac{DO} flux estimates as hourly rates of \ac{DO} change were first averaged during day and night periods for each 24 hour `metabolic day' in the time series. The `metabolic day' was considered the period between sunsets on two adjacent calendar days.  Respiration was hourly \ac{DO} flux during night hours and net production was hourly \ac{DO} flux during day hours.  Total respiration (Rt) rates were assumed constant during day and night such that daily rates were calculated as the average \ac{DO} flux during night hours multiplied by 24. Daily gross production (Pg) was the average \ac{DO} flux during day hours minus the average daily respiration, multiplied by total sunlight time.  Net ecosystem metabolism was gross production (positive) plus total respiration (negative).  Finally, volumetric rates were converted to depth-integrated (areal) estimates (g O$_2$ m$^{-2}$ d$^{-1}$, \cref{tab:case_att}) by multiplying by the mean water-column depth at the site.  Site depth was taken from the water quality sensor and a half-meter was also added to account for approximate placement slightly off of the bottom.

@@

For all case studies, weighted regression was generally successful in reducing the variation in the \ac{DO} time series that was presumably caused by physical advection.  In particular, filtering reduced the frequency of anomalous metabolism estimates, such as negative production and positive respiration \citep{Needoba12}.  It is important to recognize, however, that `normal' estimates (positive production and negative respiration) may still include significant bias due to physical advection.  For example, a large decrease in \ac{DO} at night may appear `normal' even if inflated by a tidal effect.  In some cases, apparent tidal effects could reflect real ecosystem processes. For example, \citet{Nidzieko14} observed that net metabolism at Elkhorn Slough was strongly heterotrophic during spring tides that occurred at nighttime because inundation of salt marshes during the night followed by draining during a falling tide during the day lead to larger respiration values.  Resolving these patterns may simply require better data.  For the weighted regression approach, sustained periods of synchrony between the solar and tidal cycles presents a problem because collinearity between the two diminishes performance.  Moreover, \ac{DO} time series may reflect changes in metabolism associated with tides that are true biological effects, rather than an artifact of advection.  For example, respiration may be higher on falling tides as compared to rising tides if outwelling substrates on the falling tide support increased metabolism \citep{Sasaki09}.  These processes could contribute to interday variation in metabolic rates, although our results suggested that this variation may be reduced after removing tidal effects (\cref{fig:case_ex}).  In many cases, the interday variation observed in metabolism using the unfiltered data was as large or larger than the long-term seasonal variation, indicating substantial effects of physical advection.  Therefore, we propose that the current analysis challenges the popular belief that substantial interday variation in metabolism is the norm.  This hypothesis should be further investigated and the method proposed herein could be used to better understand temporal variation of metabolism at different scales.
